A Formalization of Complete Discrete Valuation Rings and Local Fields

### 2023-12-12 HAL/CPP_paper_v2.tex
added isabelle

The field of $p$-adic numbers was later formalized in Isabelle/HOL in~\cite{padicIsa}, where we also find a formalization of formal Puiseux series \cite{PuiseuxSeriesIsa}, which are a generalization of Laurent series. To the best of our knowledge, in none of these formalizations the algebraic structure of those rings as DVR's, nor the comparison between the adic and the metric topology, is discussed.

@@

\paragraph{The \code{normed_field} and the \code{valued} instances}
\mi{Recall from~\S\ref{subsec:extensions}} that to each discrete valuation $v$ on a field $K$ we can associate a nonarchimedean multiplicative norm $\lvert\cdot\rvert_K$.

\fil{When there exists a \st{If $v$ is the }}preferred discrete valuation \fil{$v$} on $K$, we often register it as a \code{valued} instance:
\begin{lstlisting}
    {K : Type*} [field K] [hv : valued K $\wzmZZ$] 
\end{lstlisting}
In this situation, the norm $\lvert\cdot\rvert_K$ associated to $v$ would be the preferred nontrivial norm on $K$ , up to rescaling, and we would like to record a corresponding \code{normed_field} instance on $K$:
\begin{lstlisting}
    {K : Type*} [normed_field K $\wzmZZ$] 
\end{lstlisting}
Note that the datum of a field is embedded into the definition of the \code{normed_field} class, while the class \code{valued} takes the field structure as an argument. It follows that declaring a \code{normed_field} instance on every field that carries a \code{valued} one leads to a loop in the typeclass inference system\footnote{\lean[4] can detect these kinds of simple loops, so this should not be an issue once our project has been ported to \lean[4].}. We \mi{\st{propose} discuss below} a concrete example to illustrate this problem, for which we studied
a trace of the instance search using the option \code{trace.class_instances} to track the typeclass inference process.

Suppose that we had declared \code{discretely_normed_field} as an instance instead of as a definition, so that every discretely valued field
\begin{lstlisting}
(K : Type*) [field K] [hv : valued K $\wzmZZ$]
  [is_discrete hv.v]
\end{lstlisting}
would automatically inherit a \code{normed_field} instance.
Under this assumption, the typeclass inference system would get into an infinite loop when searching for an instance of
%}
\code{valuation_ring hv.v.valuation_subring.to_subring,}\relax 
because one of the first results that Lean tries to apply is \code{valuation_ring.of_field}, that automatically infers a valuation ring structure on every field. In turn, this reduces the problem at hand to finding an instance of \code{field hv.v.valuation_subring.to_subring.}
This leads to a dead end (since the valuation subring is not a field); however, while trying to find this instance, the inference system finds \code{normed_field.to_field}\relax, and hence it starts to search for an instance of
\code{normed_field hv.v.valuation_subring.to_subring}, for which it will try to apply the definition \code{discretely_normed_field}, \mi{\st{which} that} as input requires a \code{field}\relax instance on the valuation subring which recreates the problem discussed above, resulting in the system getting stuck in an infinite loop.

>>>>

\fil{
The $p$-adic numbers were later formalized in Isabelle/HOL, whose main library also contains a formalization of formal Puiseux series. The formalization of $\ZZ_p$ in~\cite{padicIntIsa} follows the classical path and is quite complete: both the definition as inverse limit and as completion of $\ZZ$ with respect to the $p$-adic valuation are formalized, together with a proof of their equivalence and of Hensel's lemma. The basic results of the topological properties of $\ZZ_p$ can also be found \ibid, and the paper~\cite{padicIsa} contains deeper results: the field $\QQ_p$ is defined as the fraction field of the ring $\ZZ_p$ formalized in the previous work, and it is endowed with both a valuation and a norm extending the previous ones on $\ZZ_p$. Again, the main topological properties both of $\QQ_p$ and of $\QQ_p^n$ are studied. Nevertheless, neither the structure of $\ZZ_p$ as a DVR, or as a Dedekind domain, are studied; also, there is no treatment of finite extensions of $\QQ_p$ or of localization results.

The work~\cite{PuiseuxSeriesIsa} presents the formalization of the ring $\puiseuxseries$ of Puiseux series (over any commutative ring $R$), that are a generalization of Laurent series. Both $\powerseries[R]$ and $\laurentseries[R]$ are defined and studied \ibid, and are endowed with a valuation and a norm, whose basic properties are formalized and extended to
\[
\puiseuxseries=\bigcup_{d\geq 1}R(\!(X^{1/d})\!).
\]
The main focus of the paper is the formalization of the Newton--Puiseux' theorem stating that whenever $C$ is an algebraically closed field of characteristic $0$, the same holds for $\puiseuxseries[C]$; as a consequence, little algebraic properties both of $\laurentseries[\FF]$ and of $\puiseuxseries[\FF]$ can be found. In particular, no formalization of the DVR structure of $\powerseries[\FF]$ or of field extensions of $\laurentseries[\FF]$ can be found \ibid.

\st{To the best of our knowledge, in none of these formalizations the algebraic structure of those rings as DVR's, nor the comparison between the adic and the metric topology, is discussed.}}

@@

\paragraph{The \code{normed_field} and the \code{valued} instances}
Recall from~\S\ref{subsec:extensions} that to each discrete valuation $v$ on a field $K$ we can associate a nonarchimedean multiplicative norm $\lvert\cdot\rvert_K$.

\fil{When there exists a \st{If $v$ is the }}preferred discrete valuation \fil{$v$} on $K$, we often register it as a \code{valued} instance:
\begin{lstlisting}
    {K : Type*} [field K] [hv : valued K $\wzmZZ$] 
\end{lstlisting}
In this situation, the norm $\lvert\cdot\rvert_K$ associated to $v$ would be the preferred nontrivial norm on $K$ , up to rescaling, and we would like to record a corresponding \code{normed_field} instance on $K$:
\begin{lstlisting}
    {K : Type*} [normed_field K $\wzmZZ$] 
\end{lstlisting}
Note that the datum of a field is embedded into the definition of the \code{normed_field} class, while the class \code{valued} takes the field structure as an argument. It follows that declaring a \code{normed_field} instance on every field that carries a \code{valued} one leads to a loop in the typeclass inference system\footnote{\lean[4] can detect these kinds of simple loops, so this should not be an issue once our project has been ported to \lean[4].}. We discuss below a concrete example to illustrate this problem, for which we studied
a trace of the instance search using the option \code{trace.class_instances} to track the typeclass inference process.

Suppose that we had declared \code{discretely_normed_field} as an instance instead of as a definition, so that every discretely valued field
\begin{lstlisting}
(K : Type*) [field K] [hv : valued K $\wzmZZ$]
  [is_discrete hv.v]
\end{lstlisting}
would automatically inherit a \code{normed_field} instance.
Under this assumption, the typeclass inference system would get into an infinite loop when searching for an instance of
%}
\code{valuation_ring hv.v.valuation_subring.to_subring,}\relax 
because one of the first results that Lean tries to apply is \code{valuation_ring.of_field}, that automatically infers a valuation ring structure on every field. In turn, this reduces the problem at hand to finding an instance of \code{field hv.v.valuation_subring.to_subring.}
This leads to a dead end (since the valuation subring is not a field); however, while trying to find this instance, the inference system finds \code{normed_field.to_field}\relax, and hence it starts to search for an instance of
\code{normed_field hv.v.valuation_subring.to_subring}, for which it will try to apply the definition \code{discretely_normed_field}, that as input requires a \code{field} instance on the valuation subring which recreates the problem discussed above, resulting in the system getting stuck in an infinite loop.
